Synthesising Concise Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-08 papers/termination/paper.tex
Compress algorithms to save space

It is clear that this loop has many non-terminating executions,
e.g. the execution where \lstinline!nondet()! always returns 0.
However it is equally clear that each state has a successor
that exits the loop, i.e.~when \lstinline|nondet()| returns
the value currently stored in \lstinline|x|.  So this loop
has an open recurrence set, but no closed recurrence set
and hence we cannot give a proof of its non-termination
with formula {\bf [CNT]} and instead must use formula {\bf [SNT]}.

@@

Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, as shown in Algorithm~\ref{alg:termination}.

\begin{algorithm}
 \caption{Bit-vector termination
 \label{alg:termination}}
 %\vspace{-1.5em}
 \begin{algorithmic}[1]
 \Statex
\Function{terminates}{$L$}
  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
  \Statex  \Comment{Formula {\bf[GT]}:}
  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
    \State \Return{$L$ terminates}
  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
    \State \Return{$L$ does not terminate}
  \EndIf
\EndFunction
 \end{algorithmic}
\end{algorithm}

Since $L$ either terminates or does not terminate, this formula is
satisfiable in the synthesis fragment.  A solution to the formula would
include witnesses $P_N$ and $P_T$, which are putative proofs of
non-termination and termination, respectively.  Exactly one of these will be
a genuine proof, so we can check first one and then the other.

@@

\begin{algorithm*}
 \caption{Abstract refinement algorithm
 \label{alg:cegis}}
 \vspace{-1.5em}
 \begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 \end{multicols}
\end{algorithm*}

@@

\begin{figure}
{\small
\begin{center}
\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

@@

\end{center}
}
 \caption{The language $\mathcal{L}$}
 \label{fig:l-language}
\end{figure}

>>>>

It is clear that this loop has many non-terminating executions,
e.g. the execution where \lstinline!nondet()! always returns 0.
However each state has a successor
that exits the loop, i.e.~when \lstinline|nondet()| returns
the value currently stored in \lstinline|x|.  So this loop
has an open recurrence set, but no closed recurrence set
and hence we cannot give a proof of its non-termination
with {\bf [CNT]} and instead must use {\bf [SNT]}.

@@

Since $L$ either terminates or does not terminate, this formula is a tautology in the synthesis fragment.
A solution to the formula would include witnesses $P_N$ and $P_T$, which are putative proofs of non-termination
and termination respectively.  Exactly one of these will be a genuine proof, so we can check
first one and then the other, as shown in Algorithm~\ref{alg:termination}:

\begin{algorithm}
{\small
 \caption{Bit-vector termination
 \label{alg:termination}}
 \vspace{-.5em}
 \begin{algorithmic}[1]
 \Statex
\Function{terminates}{$L$}
  %\Let{termination-spec}{$\exists W, R. \forall x, x' . \phi(W, R, x, x')$} \Comment{Formula {\bf [CT]}}
  %\Let{nonterm-spec}{$\exists N, C, x_0, \forall x. \psi(N, C, x_0, x)$} \Comment{Formula {\bf [SNT]}}
  \Statex  \Comment{Formula {\bf[GT]}:}
  \Let{spec}{$\exists P_T, P_N. \phi(P_T, x, x') \vee \psi(P_N, y)$}
  \Let{$(P_T, P_N)$}{{\sc Synthesise}(spec)}
  \If{decide($\exists x, x' . \lnot \phi(P_T, x, x'))$ = unsat}
    \State \Return{$L$ terminates}
  \ElsIf{decide($\exists x . \lnot \psi(P_N, x))$ = unsat}
    \State \Return{$L$ does not terminate}
  \EndIf
\EndFunction
 \end{algorithmic}
 }
\end{algorithm}

@@

%\begin{figure}
\begin{algorithm}
{\small
 \caption{Abstract refinement algorithm
 \label{alg:cegis}}
 %\vspace{-1.7em}
 %\begin{multicols}{2}
 \begin{algorithmic}[1]
 \Statex
\Function{synth}{{\sc inputs}, $\sigma$}
  \Let{$(i_1, \ldots, i_N)$}{{\sc inputs}}
  \Let{query}{$\exists P, x_0 . \forall i \in \text{\sc inputs} . \sigma(P, x_0, i)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{unsatisfiable}
  \EndIf
\EndFunction
\Statex
\Function{verif}{$P, x_0, \sigma$}
  \Let{query}{$\exists x . \lnot \sigma(P, x_0, x)$}
  \Let{result}{decide(query)}
  \If{result.satisfiable}
    \State \Return{result.model}
  \Else
    \State \Return{valid}
  \EndIf
\EndFunction
\columnbreak
\Statex
\Function{Synthesise}{$\sigma$}
  \Let{{\sc inputs}}{$\emptyset$}
  \Loop
    \Let{candidate}{\Call{synth}{{\sc inputs}}}
    \If{candidate = UNSAT}
      \State \Return{unsatisfiable}
    \EndIf
    \Let{res}{\Call{verif}{candidate}}
    \If{res = valid}
      \State \Return{candidate}
    \Else
      \Let{{\sc inputs}}{{\sc inputs} $\cup$ res}
    \EndIf
  \EndLoop
\EndFunction
 \end{algorithmic}
 %\end{multicols}
 }
\end{algorithm}
%\end{figure}

@@

\begin{figure}
\begin{center}
{\small

\setlength{\tabcolsep}{16pt}
Integer arithmetic instructions:

@@

}
\end{center}

 \caption{The language $\mathcal{L}$}
 \label{fig:l-language}
\end{figure}
